\documentclass[FCD_GNN.tex]{subfiles}

\begin{document}
\chapter{Discussion}
\label{chapter:Discussion}

The experiments conducted in this thesis demonstrate that incorporating textual
information into surface-based GNN architectures can substantially improve the
detection of FCD type~II.
At the same time, the results highlight important trade-offs that must be
considered when designing multimodal models, particularly with respect to the
balance between sensitivity and specificity.

\subsection*{Impact of textual guidance in the basic experiments}
The basic experiments show that the form and granularity of textual input play a
crucial role in multimodal lesion detection.
Among all configurations, \textbf{Vision+LM\_hemi+lobe\_regions} achieves the most
balanced performance while providing the highest sensitivity.
Incorporating hemisphere- and lobe-region information
constrains the search space, leading to improved lesion localization.

The \textbf{Vision+LM\_hemi+lobe+PubMedBERT} configuration yields the highest
overall sensitivity and specificity across both cohorts.
The concise prompt without irrelevant context, combined with PubMedBERT’s domain-specific pretraining, 
improves detection performance while only slightly reducing segmentation metrics such as Dice, IoU, PPV.

In contrast, the \textbf{Vision+LM\_full\_desc} variant underperforms relative to
the hemi+lobe-based prompts.
This suggests that long atlas-style descriptions introduce redundancy and noise,
whereas shorter, more targeted prompts provide a clearer guidance signal to the model.

\subsection*{Role of GNN-based aggregation}
Ahe analysis of experiments linking MELD features to a GNN block reveals a
nuanced picture.
Aggregating information from neighboring nodes improves the model’s ability to
detect lesions, as reflected by increased sensitivity and lesion-level coverage.
This indicates that message passing can amplify subtle abnormal patterns that may
be missed by purely local representations.

However, this benefit comes at the cost of reduced specificity, manifested as an
increased false-positive rate on healthy controls.
This is an important limitation for clinical use, since our evaluation does not include other epilepsy-related abnormalities or non-FCD lesion types.
Therefore, we cannot assess how often the model would spuriously highlight regions in patients with different pathologies.

Importantly, increasing the depth of GNN propagation does not lead to monotonic
performance improvements.
For higher-dimensional and sparser MELD feature stages, neighborhood aggregation
stabilizes representations and enriches contextual information.
In contrast, for lower-dimensional and denser stages (e.g., stages~6--7),
additional message passing induces oversmoothing, blurring spatial distinctions
and degrading performance.

\subsection*{Influence of mixed textual input and warm-up training}
Experiments with mixed textual descriptions demonstrate that even coarse prompts,
such as hemisphere- or lobe-only descriptions, can meaningfully improve lesion
detection compared to purely vision-based baselines.
Across configurations, gains in sensitivity typically come with a trade-off in specificity, i.e., an increased number of false positive clusters.
Overall, reduced--text prompts (e.g., hemisphere + lobe) often provide a more favorable balance between detecting lesions and limiting false positives, which may be preferable for clinical use.

The warm-up experiments further indicate that freezing the visual backbone during
the initial training epochs can be beneficial.
A plausible explanation is that this strategy reduces sensitivity to prompt
variability and mildly incorrect descriptions.
This behavior can be attributed to three factors: (i) even imperfect prompts
convey coarse structural information that narrows the search space, (ii) the
decoder exhibits robustness to mild text--image mismatch due to pretraining, and
(iii) training with slightly noisy prompts acts as a regularizer, discouraging
overfitting to specific wording and promoting more conservative predictions.

\subsection*{Observations from appendix analyses}
The text distribution analysis shows that conditioning on fine-grained region
names tends to induce overfitting due to the highly imbalanced and long-tailed
label distribution.
Frequent regions dominate the learning signal, while rare categories lack
sufficient variability for robust generalization.

The semantic similarity analysis further highlights the importance of the text
encoder.
Biomedical language models pretrained on domain-specific corpora (e.g.,
PubMedBERT, RadBERT, BlueBERT) exhibit higher semantic coherence between
anatomically related brain regions.
This suggests that semantically aligned text embeddings facilitate more effective
multimodal fusion and improve both interpretability and performance.

\subsection*{General Observations}
Although the observed improvements in sensitivity and lesion-level detection are
consistent across experiments, the results should be interpreted as a
methodological proof of concept demonstrating the benefit of fusing visual and
textual information, rather than as a clinically deployable solution.
In this work, textual input is limited to structured, atlas-derived descriptions
with explicit anatomical labels.
Importantly, the performance gains cannot be attributed to the text providing
explicit supervision about lesion presence: for healthy subjects, the text
indicates ``no lesion detected'', whereas for patients with FCD, including the
``wrong text'' setups, the text only provides coarse spatial information (e.g.,
hemisphere or lobe) and does not encode whether a lesion is present.
The main effect of textual guidance is therefore a spatial regularization of the
search space, which can improve sensitivity but may also increase false positives
when the anatomical prior is inaccurate.


\subsection*{Limitations and future directions}
Several limitations must be acknowledged. 
While the MELD surface features are ComBat-harmonized to reduce inter-site effects, residual variability may remain due to differences in acquisition protocols, preprocessing, and lesion annotation practices across centers.
Furthermore, the MELD cohort predominantly consists of histologically confirmed type II FCD cases, and it therefore remains unclear whether the same conclusions extend to other FCD subtypes.
Finally, the model relies on atlas-based text generated automatically; incorporating free-form radiology reports could yield a more realistic estimate of performance under clinical conditions, where textual descriptions are not always precise.

This work does not explore the effect of unfreezing different numbers of the
language model layers or varying the number of cross--modal connections to the
GuideDecoder.
However, previous work (Huemann et al., 2024~\cite{Huemann2024ConTEXTualNet})
shows that selectively unfreezing specific layers of a language model and
fine-tuning them for a segmentation task can yield consistent performance gains.
These aspects remain for future investigation.

Another conceptual direction concerns the representation of anatomical descriptions. 
In this thesis, atlas information was encoded via textual prompts and processed by a pretrained language model. 
An alternative would be to express the same information as a numerical feature vector with one dimension per cortical region. 
In such a vector, the percentages mentioned in the textual description (e.g., 52.34\% Right Frontal Orbital Cortex; 26.06\% Right Frontal Pole) would directly populate the corresponding entries, while all other regions would be set to zero. 
For coarse labels such as lobe-only or hemisphere-only descriptions, the corresponding weight could be uniformly distributed across all constituent regions.

While this vector--based formulation is simple, interpretable, and avoids the computational overhead of a pretrained LM, it is also less expressive in its basic form. 
When used without an explicitly defined structure, a model trained solely on such numerical representations treats each vector dimension as an independent feature and does not explicitly encode anatomical relationships, such as lobar or hemispheric grouping, spatial adjacency, or hierarchical organization. 
Although these relationships could in principle be incorporated through carefully designed numerical encodings or structured representations, this requires additional modeling assumptions and design choices.
In contrast, language models implicitly encode semantic relationships between anatomical terms, including regional similarity and hierarchical structure, and naturally handle coarse, uncertain, or partially specified descriptions. 
This makes LM--based representations particularly suitable for prototyping multimodal fusion and exploring how textual guidance can complement visual features. 
Accordingly, while numerical vector representations remain a sustainable baseline, their relative performance compared to LM-based approaches depends on how explicitly anatomical structure and uncertainty are modeled, and remains an open question for future work.


Overall, the results indicate that textual guidance is a promising direction for
FCD detection.
However, the balance between sensitivity and precision remains a critical design
choice that must be adapted to the clinical context and intended use case.

% \item How does fine-tuning different numbers of layers in a pre-trained \ac{llm} affect model performance?
% \item Performed extensive experiments to evaluate different design choices, such as the number of connections in the \ac{gnn} blocks and the use of various textual branches in the GuideDecoder, to improve segmentation results.
% Chapter~\ref{chapter:Experiments} details the experimental setup (losses, metrics, and implementation specifics) and reports the main results: baseline comparisons against MELD, the effect of linking different numbers of MELD feature stages to the \acs{gnn} block, the influence of (partially) unfreezing RadBERT and using text guidance in the decoder, and ablations on the number of text–GuideDecoder connections; where relevant, we also compare pre- and post-fine-tuning performance.
\end{document}
