\documentclass[FCD_GNN.tex]{subfiles}

\begin{document}
\chapter{Related Works}
\label{chapter:Related_works}

This section firstly reviews the transformer architecture, which is a fundamental component of modern vision models and large language models (LLMs).
Then it introduces graph neural networks (GNNs) and LLMs, followed by an overview of existing approaches to \ac{fcd} detection and text-guided medical image segmentation for tumor detection tasks, with a focus on their fusion strategies, and limitations.

\section{Transformer Architecture}
The transformer architecture, introduced in the seminal work by Vaswani et al.~\cite{Vaswani2017Attention}, is based entirely on attention mechanisms and does not rely on recurrence or convolution. 
Its core component is the \textbf{\textit{self-attention layer}}, which models long-range interactions between elements of a sequence. 
Given input embeddings $X$, the model computes query, key, and value matrices $Q = XW^Q$, $K = XW^K$, and $V = XW^V$, where $W^Q$, $W^K$, and $W^V$ are learnable projections respectively. 
The self-attention operation is defined as:
\begin{equation}
    \mathrm{Attention}(Q, K, V) = 
    \mathrm{softmax}\left(\frac{QK^{\top}}{\sqrt{d_k}}\right)V,
\end{equation}
where $d_k$ is the dimensionality of the key vectors.

To enhance representational capacity, transformers employ 
\textbf{\textit{multi-head attention}}, which performs several attention operations
in parallel:
\begin{equation}
    \mathrm{MultiHead}(Q,K,V) = 
    \mathrm{Concat}(\mathrm{head}_1,\dots,\mathrm{head}_h)\, W^O,
\end{equation}
where each head computes self-attention using its own learnable
projections:
\begin{equation}
    \mathrm{head}_i = 
    \mathrm{Attention}(QW_i^{Q},\, KW_i^{K},\, VW_i^{V}),
\end{equation}
with parameter matrices 
$W_i^{Q} \in \mathbb{R}^{d_{\text{model}} \times d_k}$,
$W_i^{K} \in \mathbb{R}^{d_{\text{model}} \times d_k}$, and 
$W_i^{V} \in \mathbb{R}^{d_{\text{model}} \times d_v}$.
The output projection 
\begin{equation}
    W^O \in \mathbb{R}^{h d_v \times d_{\text{model}}},
\end{equation}
mixes information across all $h$ heads and restores the dimensionality back to 
$d_{\text{model}}$.


Originally developed for natural language processing, transformer-based models have since been adapted to computer vision (e.g., ViT~\cite{Dosovitskiy2020ViT}, Swin Transformer~\cite{Liu2021Swin}) and to a variety of multimodal learning settings.
In text-guided medical image segmentation, transformers are commonly used either as text encoders or as fusion modules, where cross-attention integrates semantic cues from textual descriptions with spatial image representations.
This capacity to incorporate global contextual information makes transformers particularly suitable for multimodal medical tasks, although their performance often depends on the availability of sufficiently large and diverse training datasets.

\section{Graph Neural Networks}
Graph Neural Networks (GNNs)~\cite{Scarselli2008GNN} provide a principled framework for learning on
non-Euclidean data such as meshes, surfaces, and general irregular structures.
In contrast to convolutional architectures that operate on grids, GNNs propagate
information along graph edges, enabling feature integration across anatomically
or structurally meaningful neighborhoods. Most modern GNNs follow the 
\textit{message-passing} formulation~\cite{Gilmer2017MessagePassing}, where node 
representations are iteratively updated by aggregating features from adjacent 
nodes. For a node $v$ with neighborhood $\mathcal{N}(v)$ and current features 
$h_v$, the generic message-passing layer is expressed as:
\begin{equation}
    m_v = \mathrm{AGG}\bigl(\{\, \phi(h_v, h_u, e_{uv}) 
    : u \in \mathcal{N}(v) \,\}\bigr),
\end{equation}
\begin{equation}
    h_v' = \gamma(h_v, m_v),
\end{equation}
where $\phi$ computes messages from neighbors, 
$\mathrm{AGG}$ is a permutation-invariant aggregation function 
(e.g., mean, sum, max), and $\gamma$ is an update function 
(often implemented as an MLP). 
Different GNN architectures instantiate these components in distinct ways. 
For example, Graph Convolutional Networks (GCN)~\cite{Kipf2017GCN} use a 
normalized mean aggregator,
while GraphSAGE~\cite{Hamilton2017GraphSAGE} generalizes aggregation through 
mean, pooling, or LSTM-based operators, enabling inductive learning on unseen 
graphs.

Hierarchical GNN architectures extend this idea to multi-resolution settings,
analogous to U-Net models~\cite{Ronneberger2015UNet} in Euclidean domains and
graph U-Net architectures~\cite{Gao2019GraphUNet}. 
They downsample and upsample graphs using pooling and unpooling operators, which
enable coarse-to-fine feature extraction on complex surfaces. 
This paradigm is particularly relevant for cortical-surface analysis, where the
brain can be represented as a multi-resolution mesh.  

Building on these principles, the MELD project~\cite{Ripart2025MELD} applies a
GNN-U-Net architecture to multi-resolution cortical graphs and demonstrates that
graph-based convolutions are effective for detecting subtle morphological
abnormalities associated with \ac{fcd}.  

In this thesis, we build upon these principles to integrate cortical surface
features with textual clinical information, enabling multimodal segmentation
within a unified GNN-based framework.


\section{Large Language Models} 
\ac{llm}s are typically based on transformer architectures and are trained on large text corpora to learn contextual semantic representations.

Given a sequence of token embeddings $X = (x_1,\dots,x_T)$, an LLM encoder 
produces hidden states
\begin{equation}
    H = (h_1, \dots, h_T),
\end{equation}
where each vector $h_i$ is a context-dependent representation of token $x_i$.
This means that $h_i$ is computed not only from the embedding of $x_i$ itself 
but also from information coming from other tokens in the sequence, which is 
achieved by the self-attention mechanism.

LLMs are typically trained using one of two objectives.
In \textit{causal} language modeling (used in GPT-style models), the model 
predicts the next token based on the previously observed ones:
\begin{equation}
    p(x_{i+1} \mid x_1,\dots,x_i) = \mathrm{softmax}(W^O h_i).
\end{equation}
In \textit{masked} language modeling (MLM), employed by BERT-style encoders, 
randomly selected tokens are masked and reconstructed from the surrounding 
context:
\begin{equation}
    p(x_i \mid X_{\setminus i}) = \mathrm{softmax}(W^O h_i),
\end{equation}
where $X_{\setminus i}$ denotes the input sequence with token $x_i$ masked.

LLMs trained with these objectives produce rich, high-level semantic embeddings 
that capture lexical, syntactic, and domain-specific patterns in text~\cite{devlin2019bert}. 
In medical imaging, domain-adapted LLMs such as BioBERT~\cite{Lee2019BioBERT} 
provide contextual text embeddings that can guide segmentation models through 
cross-attention or contrastive alignment~\cite{Radford2021CLIP}, improving robustness and interpretability 
in multimodal medical applications.

\section{State-of-the-Art Methods}
Building on early vision-based approaches in medical imaging, initial automated methods for FCD detection relied on convolutional neural networks operating directly on MRI data.
Dev et al.~\cite{dev2019automatic_fcd} proposed a fully convolutional neural network with an encoder--decoder structure and skip connections, conceptually similar to a U-Net, for the automatic detection and localization of \ac{fcd} from volumetric MRI.
Their approach performs voxel--wise segmentation on multimodal MRI data, enabling direct lesion identification without manual feature engineering.
However, the method was trained on a relatively small, single--center dataset, which may limit its robustness and generalizability across different populations and imaging protocols.

Similarly, House et al.~\cite{house2021automated_fcd} introduced a clinically integrated 3D convolutional neural network for the automated detection and segmentation of \ac{fcd} lesions from routine T1--weighted and FLAIR MRI sequences.
The proposed architecture employs a shared encoder that extracts latent representations from the MRI data, followed by two decoders: one dedicated to FCD segmentation and a variational decoder tasked with reconstructing the input to model normal brain anatomy.
This joint training strategy regularizes the encoder and improves segmentation performance, particularly in settings with limited labeled data.
A key contribution of this work is its prospective validation within a real--world clinical workflow.
Nevertheless, the model exhibits substantially lower detection performance for subtle or non-sclerotic \ac{fcd} lesions compared to more conspicuous cases, suggesting a bias toward visually pronounced lesion patterns and limited generalization across the full pathological spectrum.

Despite their architectural differences, both approaches operate purely in volumetric image space and rely primarily on intensity-based MRI features.
As a result, they remain sensitive to noise, imaging artifacts, and inter-scanner variability, and do not explicitly model cortical morphometric characteristics such as cortical thickness, curvature, or gray--white matter blurring, which are key markers of FCD.
These limitations make subtle lesions particularly difficult to detect and motivate the use of surface-based representations that better capture cortical anatomy.

A more recent and substantial advancement is represented by the \ac{meld} Graph model~\cite{Ripart2025MELD}, 
which models the cerebral cortex as a multi-resolution graph and applies a \ac{gnn}-U-Net architecture for lesion segmentation.
Trained on the largest multi-center FCD dataset available to date, the MELD model represents the current state of the art in surface-based FCD detection and demonstrates competitive performance across multiple evaluation settings, while substantially improving robustness to inter--site variability between imaging centers.
Owing to its strong performance to date and its surface-based representation of cortical features, the MELD Graph model serves as a well-established vision backbone for the present study.

Recent progress in multimodal learning suggests that incorporating textual information alongside visual features can improve performance in data--limited medical imaging tasks.
In particular, Ariadne’s Thread~\cite{Zhong2023Ariadne} demonstrated that integrating textual descriptions via a cross--attention mechanism leads to consistent performance gains for tumor detection, even when only a small number of annotated training samples is available.
This finding is especially relevant for FCD detection, where annotated data are scarce and lesion appearance is highly heterogeneous.
Motivated by these results, we investigate whether similar benefits can be achieved by augmenting a strong surface-based vision model with textual anatomical information.
% Specifically, we adopt the pretrained MELD Graph model as the visual backbone and extend it with atlas-derived textual descriptions within an architecture inspired by Ariadne’s Thread.

% Early work on \ac{fcd} detection focused exclusively on vision-based methods~\cite{Durgam2025LungCancer, Abdusalomov2023BrainTumor, Sharma2025KidneyCT, Mehmood2025BreastCancer}. 
% It is noteworthy that one of the newest models, the \ac{meld} Graph~\cite{Ripart2025MELD}, represents the surface of the cerebral cortex as a graph with multiple resolutions and applies a \ac{gnn}-U-Net to segment lesions. It provides high 
% sensitivity and specificity by identifying characteristic peaks (more than 20\% in saliency) and determining calibration reliability using the expected 
% calibration error. However, it has not been tested on patients with multiple \ac{fcd}s, it lacks cross-attentional mechanisms for more complete integration of 
% features, and it does not integrate textual clinical information or evaluate zero-shot generalization across FCD subtypes.

% A second line of research explores language-guided segmentation by embedding text semantics into the segmentation pipeline. 
% Early methods~\cite{Tomar2022TGANet, Li2023LViT} simply tokenized text and merged embeddings with image features via attention, 
% but struggled to capture high-level semantics. 
% More recent approaches~\cite{Lee2023TextGuided, Zhong2023Ariadne} employ deep pretrained text encoders (e.g., CXR-BERT~\cite{CXR-BERT}) fused with ConvNeXt-Tiny~\cite{Liu2022ConvNeXt} image features. 
% In~\cite{Li2024LGMSeg}, the authors introduce two novel modules: the Target-sensitive Semantic Distance Module (TSDM), 
% which computes contrastive distances between segmentation masks to focus on disease-related regions, 
% and the Language-guided Target Enhancement Module (LTEM), which applies cross-attention to reinforce critical image areas. 
% A bi-directional contrastive loss (averaged cross-entropy in both ``image $\rightarrow$ text'' and ``text $\rightarrow$ image'' directions) yields more fine-grained guidance 
% and enables models trained on 25\% of the data to outperform single-modal baselines trained on full datasets. 
% However, these improvements come at the cost of higher architectural complexity and a strong reliance on high-quality textual annotations, 
% which may limit robustness and generalizability in real-world clinical settings.

% To address unpaired multi-modal data, MulModSeg~\cite{Li2025Mulmodseg} conditions text embeddings on imaging modality using frozen CLIP~\cite{Radford2021CLIP}/BioBERT~\cite{Lee2019BioBERT}/Med-CLIP~\cite{Wang2022MedCLIP} encoders 
% combined with medical prompts, and alternates training between vision and text branches (3D-UNet\cite{Cicek2016UNet3D} or SwinUNETR~\cite{Hatamizadeh2022SwinUNETR}). This scheme improves generalization 
% without requiring paired \ac{ct}/MR scans. Experiments show that varying the \ac{ct}:MR ratio in training data shifts performance, underscoring the importance 
% of balanced modality representation. However, effectiveness depends critically on prompt template design and alternating-training convergence.

% Finally, weakly supervised methods such as SimTxtSeg~\cite{Xie2024SimTxtSeg} leverage simple text cues 
% (e.g., ``lesion in the left hemisphere'') to guide segmentation without requiring pixel-level annotations. 
% SimTxtSeg was evaluated using standard segmentation metrics, including Dice similarity coefficient (DSC), 
% Intersection over Union (IoU), Positive Predictive Value (PPV), Normalized Surface Dice (NSD), and the 
% 95\% Hausdorff Distance (HD95), alongside cross-entropy loss. These results demonstrate that even coarse 
% textual hints can substantially improve mask quality over vision-only baselines. However, such approaches 
% typically rely on consistent, high-quality text labels and have not yet been validated on rare pathologies 
% such as \ac{fcd}.

% Building on these insights, we draw inspiration from Ariadne’s Thread~\cite{Zhong2023Ariadne}, which employs 
% lightweight text prompts and a GuideDecoder to segment infected regions in chest X-rays. Remarkably, this method 
% achieves over a 6\% Dice improvement compared to unimodal baselines while using only 10\% of the training set, 
% highlighting the potential of multimodal prompting in low-data regimes. For our FCD task, we adopt the \ac{meld} 
% Graph model as a pretrained backbone, since it was trained on the largest dataset among the methods surveyed, 
% providing a robust basis for feature extraction and downstream adaptation.

% As far as we know, no prior work has applied text-guided segmentation to the \ac{fcd} detection task, underscoring the novelty of our approach. 
% In this work, we demonstrate that incorporating textual information benefits \ac{fcd} detection. 
% To this end, we integrate atlas-derived textual descriptions with visual features in a multi-resolution GNN framework, using cross-attention fusion and contrastive alignment to improve robustness and detection accuracy across heterogeneous datasets.

\end{document}

